Hydraulically-vulnerable trees survive on deep-water access during droughts in a tropical forest

### 2021-04-09 09.0_manuscript.rmd
Typo corrections and other minor edits

Section  | Word Count  
------------- | -------------  
Summary | 200  
Main Text  | 6939  
Introduction | 832  
Methods |  3150  
Results | 855  
Discussion | 2102  
Keywords | 8

@@

\
Dry season deciduous species may escape drought exposure via leaf deciduousness, so we analysed deciduous and evergreen species separately. Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; see SI). We pooled all deciduous leaf habits together (hereafter 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.

@@

The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)**a**, **b**), including the out-of-sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)**c**) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was primarily sensitive to the Ball-Berry stomatal slope parameter (fates_leaf_BB_slope), the ELM root distribution parameter that regulates the depth of the rooting profile (fates_rootb_par), soil hydraulic conductivity (HKSAT) profile especially at depth, and the adjustment factor (HKSAT_ADJ) that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs. S4, S5). The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m with the median at `r root.zone.med` m.

\
Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Hydrological droughts in census interval 1990-1995 was marked by a prolonged dry season, while those during 2000-2005 and 2005-2010 distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S6).

@@

The best ERD model explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig. \@ref(fig:erd-iso); also see SI). This model included an effect of VPD and not LAI (Eq. \@ref(eq:M-chosen)).

@@

Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El-Niño event [@Detto.2018; @Condit.2017]. Out of the six intervals with El-Niño events ERD explained 30-40% variation in mortality rates among evergreen species during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD (p-values < 0.05 for four intervals and 0.06 for one interval; Fig. \@ref(fig:erd-mort-census-evg)). For deciduous species ERD explained 11-16% variation in mortality in four intervals with El-Niño events, but p-values were not significant (Fig. S11).

@@

We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (\@ref(eq:M-chosen), Figs. \@ref(fig:schematic), \@ref(fig:sp-erd)). Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Figs. \@ref(fig:erd-iso)). Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but lower safety (Fig. \@ref(fig:erd-hyd)). Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig. \@ref(fig:erd-mort-census-evg), S11). This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Niño events---over seven census-intervals studied in total (1982-2015). Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig. \@ref(fig:swp-droughts)). Species exposure to water-stress exponentially declined with deeper ERD (Fig. \@ref(fig:erd-time-evg)). Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018]. We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality across several El-Niño events over 35 years.

@@

If investments in stress-tolerance traits come at a cost, such insurance may not pay-off if the risk of stress is not realised. Our findings suggest that species with investment in deep roots can afford the hydraulically efficient, but risky suite of traits (Fig. \@ref(fig:erd-hyd)), because access to a reliable deep-water resource ensures that for them hydraulic risk is not realised (Figs. \@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Fig. S6).

\
Extreme, prolonged El-Niño droughts (Fig. \@ref(fig:swp-droughts)) in our study may have crossed even the greater tolerance limits of shallow-rooted species leading to their greater mortality compared to deep-rooted species (Fig. \@ref(fig:erd-mort-census-evg)). Hydraulic risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century [@Condit.2017]. Greater survival of deep-rooted species that we observed may change into the future if droughts intensify.

\
We found significant ERD-mortality relationships during extreme water-stress for evergreen but not deciduous species (Fig. \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop. This is also consistent with the observation that species distribution along local (BCI) and regional (Panama) moisture gradients is correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021]. Studies assessing coordination between leaf phenology, rooting depth and hydraulic traits simultaneously are almost absent for the tropics and warrant future consideration [@Oliviera.2021]. Our analyses for ERD-hydraulic traits relationships were limited to only seven species, but covered all species of different leaf phenologies (Fig. \@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phenologies. We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig. \@ref(fig:sp-erd)), consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021]. Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016]. Leaf phenologies at BCI are numerous and complicated and warrant research.

@@

\
Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019; @Brum.2017; @Rowland:2015]. Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010]. Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery [@Bennett:2015; @McGregor.2020]. Future studies should test the ERD-mortality relationship on a greater number of species.

\
Our result of lower mortality in deep-rooted trees contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally dry tropical forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see [@Goulden.2019; @Ivanov:2012]). In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 1 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (-0.28 MPa) for all the dry seasons and droughts during 1990-2015. Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI are major factors in the different mortality responses. At BCI deep soil layers were recharged annually (Figs. S6, \@ref(fig:swp-droughts)), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.

@@

The relationships that we identify between above- and below-ground traits, vertical profile of soil water status and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies. In the context of trait filtering models, if we used the hydraulic trait information without knowledge of their relationship to rooting depth, models would likely kill the 'risky' strategy trees in droughts, which would in fact be the opposite result from that observed in this study. A key outcome of this study is thus the relationships between hydraulic traits and ERD that could be plugged into a DGVM of BCI. To assimilate ERDs, DGVMs could vary rooting parameters such that the centroid of the species water-uptake profiles match ERDs. We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves. Albeit future studies should undertake sensitivity analyses for uncertainties involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find thus provide important insights on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015; @Mencuccini.2017], but that may provide an avenue to include stem water storage and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S9). Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from a leaf lifetime, and omits inter-annual variation. Species-level leaf-fall data from litter-traps that we use includes large within-species variability in leaf-fall timing, and thus does not capture typical tree-level deciduous seasonality. Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

>>>>

Section  | Word Count  
------------- | -------------  
Summary | 200  
Main Text  | 6934  
Introduction | 831    
Methods |  3150  
Results | 838  
Discussion | 2115  
Keywords | 8

@@

\
Dry season deciduous species may escape drought exposure via leaf deciduousness, so we analysed deciduous and evergreen species separately. Tree species on BCI are scored by expert botanists as one among four leaf-habits; evergreen, brevideciduous, facultative deciduous and obligate deciduous (@Meakem.2018; see SI). We pooled all deciduous leaf habits together (hereafter, 'deciduous' for brevity) and regressed species ERD against species $\textrm{M}_{s,t}$ for each census interval $t$ for deciduous ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous != "E"))`) and evergreen species ($n$ = `r nrow(mrate.depth.mean %>% subset(deciduous == "E"))`) separately.

@@

The 100 ELM-FATES ensemble members with the best fits to observed soil moisture dynamics captured soil moisture seasonality at multiple depths (Fig. \@ref(fig:hydro-calib)**a**, **b**), including the out-of-sample observations. These simulations also captured the dynamics in stream discharge (Fig. \@ref(fig:hydro-calib)**c**) and in evapotranspiration (Fig. \@ref(fig:hydro-calib)**d**), slightly underestimating peak discharge in wet years and slightly overestimating peak ET. The reduction in parameter range (Table S1) in the best-fit ensembles compared to the tested global ranges showed that the model calibration was primarily sensitive to the Ball-Berry stomatal slope parameter (fates_leaf_BB_slope), the ELM root distribution parameter that regulates the depth of the rooting profile (fates_rootb_par), soil hydraulic conductivity (HKSAT) profile especially at depth, and the adjustment factor (HKSAT_ADJ) that modifies soil hydraulic conductivity to account for macroporosity and direct flow paths (Figs. S4, S5). The distribution of maximum depth of soil water dynamics, and thus ecosystem root zone depth, across the 100 hydrological realisations encompassed 95% CI of `r root.zone.CI1` m - `r root.zone.CI2` m (median at `r root.zone.med` m).

\
Soil water potential dynamics $\mathrm{\Psi}_{\textrm{soil},z}$ by depth z from 0.01 through 13 m obtained from the best-fit simulations revealed that every 5-yr census interval had at least one extreme hydrological drought year, but each interval varied in terms of number of extreme years, drought intensity, drought seasonality and duration (Fig. \@ref(fig:swp-droughts)). Hydrological droughts in census interval 1990-1995 was marked by a prolonged dry season, while those during 2000-2005 and 2005-2010 distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). Across the simulation period 1990-2015, simulated $\mathrm{\Psi}_{\textrm{soil},z}$ remained above -0.5 MPa from depths 1.7 to 13 m (Fig. S6).

@@

The best ERD model (Eq. (3)) explained a large fraction of the variance in $\mathrm{\delta}$^2^H~xylem~ ($R$^2^ = `r lm.erd.iso.r2`, $p$ = `r lm.erd.iso.p`, $n$ = `r length(unique(erd.iso$sp))`, Fig. \@ref(fig:erd-iso); also see SI). This model included an effect of VPD and not LAI (Eq. \@ref(eq:M-chosen)).

@@

Among the seven census-intervals over 1982-2015 for which we analysed relationship of ERD with mortality rates, six intervals were associated with occurrence of one or more El-Niño event [@Detto.2018; @Condit.2017]. Out of the six intervals with El-Niño events ERD explained 30-40% variation in mortality rates among evergreen species during five intervals--1982-1985, 1985-1990, 1995-2000, 2000-2005 and 2005-2010--such that species mortality rates decreased with deeper ERD ($p$-values < 0.05 for four intervals and 0.06 for one interval; Fig. \@ref(fig:erd-mort-census-evg)). For deciduous species ERD explained 11-16% variation in mortality in four intervals with El-Niño events, but p-values were not significant (Fig. S11).

@@

We introduce a novel approach for estimating ERD using 25 years of tree growth, species-specific leaf vulnerability curves, modeled soil water potential profiles, and observed VPD (Eq. \@ref(eq:M-chosen), Figs. \@ref(fig:schematic), \@ref(fig:sp-erd)). Our predictions of ERD were consistent with estimates using $\mathrm{\delta}$^2^H~xylem~ (Fig. \@ref(fig:erd-iso)). Our analyses suggest that co-occurring large canopy tree species with deeper ERD were associated with higher above-ground hydraulic efficiency, but lower safety (Fig. \@ref(fig:erd-hyd)). Nevertheless, evergreen but not deciduous species with deeper ERD showed significantly lower mortality rates than shallower ERD species (Fig. \@ref(fig:erd-mort-census-evg), S11). This ERD-mortality relationship was significant in five out of six census-intervals that had experienced one or more El-Niño events---over seven census-intervals studied in total (1982-2015). Quantifying extreme droughts in the whole soil column over 1990-2015 revealed that ERD explained significant levels of mortality in periods when extreme soil droughts prolonged the dry season water-stress (Fig. \@ref(fig:swp-droughts)). Species exposure to water-stress exponentially declined with deeper ERD (Fig. \@ref(fig:erd-time-evg)). Because extreme, sustained water-stress increases risk of mortality via hydraulic-failure and/or carbon starvation, deeper ERD may mitigate drought-induced mortality by limiting exposure to water-stress [@Rowland:2015; @McDowell:2018]. We thus demonstrate, for the first time, a link between species trade-offs in above- and below-ground hydraulic traits, drought exposure through extreme hydrological droughts quantified over the whole soil-column, and large ($\ge$ 10 cm DBH) tree mortality across several El-Niño events over 35 years.

@@

If investments in stress-tolerance traits come at a cost, such insurance may not pay-off if the risk of stress is not realised. Our findings suggest that species with investment in deep roots can afford the hydraulically efficient, but risky suite of traits (Fig. \@ref(fig:erd-hyd)), because access to a reliable deep-water resource ensures that for them hydraulic risk is not realised (Fig. \@ref(fig:erd-time-evg)), whereas shallow rooted species pay the cost of hydraulic safety in terms of efficiency, adapted for an environment in which hydraulic risk is significant, as extreme droughts cause exponentially greater water-stress in shallow soil layers (Figs. Figs. \@ref(fig:erd-time-evg), S6).

\
Extreme, prolonged El-Niño droughts (Fig. \@ref(fig:swp-droughts)) in our study may have crossed even the greater tolerance limits of shallow-rooted species leading to their greater mortality compared to deep-rooted species (Fig. \@ref(fig:erd-mort-census-evg)). Hydraulic risk for shallow rooted species may have been exacerbated in our 35-yr study period in which extreme water-deficit years were more frequent than those in the last century [@Condit.2017]. Greater survival of deep-rooted species that we observed may not continue into the future if droughts intensify.

\
We found significant ERD-mortality relationships during extreme water-stress for evergreen but not deciduous species (Fig. \@ref(fig:erd-mort-census-evg), S11); consistent with the expectation that deciduous species can also avoid water-stress via leaf-drop. This is also consistent with the observation that species distributions along local (BCI) and regional (Panama) moisture gradients are correlated with $\mathrm{\Psi}$~tlp~ for evergreen but not deciduous species [@Kunert.2021]. Studies that simultaneously assess coordination between leaf phenology, rooting depth and hydraulic traits are almost absent for the tropics and warrant future consideration [@Oliviera.2021]. Our analyses for ERD-hydraulic traits relationships were limited to only seven species, but covered all species of different leaf phenologies (Fig. \@ref(fig:erd-hyd)), so here we assume that the trend in the ERD-hydraulic trait relationships holds true across all species of different leaf phenologies. We also found that the range of ERDs overlapped across leaf phenology, but evergreen species tend to have deeper ERDs (Fig. \@ref(fig:sp-erd)), consistent with observations by @Meinzer:1999 and elsewhere [@Smithmartin.2020; @Fan:2017; @Oliviera.2021]. Whether deciduous species also have more efficient and vulnerable hydraulics at BCI as is observed elsewhere remains to be studied [@Markesteijn:2010; @Markesteijn:2011; @Xu:2016; @Gleason.2016]. Leaf phenologies at BCI are numerous and complicated and warrant further research.

@@

\
Our results are consistent with recent studies that analysed rooting or water-sourcing depths versus hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see, @Brum.2019; @Brum.2017; @Rowland:2015]. Globally, large trees tend to exhibit greater growth reductions, lower post-drought resilience, and greater increases in mortality relative to their understory counterparts [@Bennett:2015; @Phillips:2010]. Our finding that deep-water access buffers drought-induced mortality in large trees is relevant for understanding drought resistance, resilience and recovery [@Bennett:2015; @McGregor.2020]. Future studies should test the ERD-mortality relationship on a greater number of species.

\
Our result of lower mortality in deep-rooted trees contrasts with the inverse model finding of @Chitra-Tarak.2018 (hereafter, CT2018) in which deeper ERD species in a South-Asian seasonally dry tropical forest had higher mortality in a rare, prolonged drought. The hydrological model in CT2018 revealed that the multi-year drought exhausted the deep soil and even bedrock water availability (also see, @Goulden.2019; @Ivanov:2012). In contrast, our hydrological modeling at BCI finds that $\mathrm{\Psi}$~soil~ for depths deeper than 2.9 m did not cross $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa) for all the dry seasons and droughts during 1990-2015 (see SI). Mean annual rainfall of 1095 mm in CT2018 compared to `r round(rain.man.stats$rain)` mm at BCI, and precipitation to potential evapotranspiration ratio of nearly one in CT2018 compared to nearly two in BCI, are major factors in the different mortality responses. At BCI deep soil layers were recharged annually (Fig. S6), whereas in CT2018 they were not. The contrast between the two studies highlights the combined role of seasonal precipitation input and site-specific hydrology in modulating mortality-risk of deep-rooted species.

@@

The relationships that we identify between above- and below-ground traits, vertical profile of soil water status and mortality rates are important for representing diversity in DGVMs, which intrinsically rely on the parameterization of contrasting life history strategies [@Scheiter:2013] and the simulation of competition between those strategies. In the context of trait filtering models, if we used the hydraulic trait information without knowledge of their relationship to rooting depth, models would likely kill the 'risky' strategy trees in droughts, which would in fact be the opposite result from that observed in this study. A key outcome of this study is thus the relationships between hydraulic traits and ERD that could be plugged into a DGVM of BCI. To assimilate ERDs, DGVMs could vary rooting parameters such that the centroid of the species water-uptake profiles match ERDs. We found that LMA and WSG were strong predictors of leaf hydraulic vulnerability curves (Fig. \@ref(fig:ab-fit)). Albeit future studies should undertake sensitivity analyses for uncertainties involved, our finding offers the promise of a greater ability to parameterize the 'hard' hydraulic traits with the abundant 'soft' trait data, thus allowing for a better representation of forest hydrodynamics. The relationships between ERD and above-ground hydraulic traits that we find thus provide important insights on how to model rooting depths and their coordination or trade-offs with other traits in order to better represent the functional diversity of tropical forests, and their trajectories into the future.

\
Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with higher frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015; @Mencuccini.2017], but that may provide an avenue to include stem water storage and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S9). Our interpretation of this result is that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate that combines leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Species-level leaf-fall data from litter-traps that we used includes large within-species variability in leaf-fall timing, and so may not have captured  tree-level seasonality in deciduousness (see SI). Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

### 2021-04-09 10.0_Supporting_Information.Rmd
Typo corrections and other minor edits

# Additional results on exposure to water-stress

Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 1.7 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).

>>>>

## Additional results on exposure to water-stress

Overall, species with ERD deeper than 2.9 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths deeper than 2.9 m remained above `r round(psi.2.9.min$median, 2)` MPa (`r round(as.numeric(psi.2.9.min$q97.5), 2)`, `r round(as.numeric(psi.2.9.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above $\mathrm{\Psi}$~crit~ of the most sensitive tree species (`r min.psi_crit` MPa).
